---
title: "Cytokinin and Auxin In Apical Dominance"
author: "Alex Lee and Katie Jensen"
date: "April 24, 2017"
output: html_document
---

##Introduction

1.) In plants, apical dominance occurs when auxin (a hormone that acidifies cell walls by causing cells to pump hydrogen atoms into the cell wall) travels rootward down the stalk, inhibiting axillary branching from occurring.  This is an indirect influence via secondary messengers. Auxin downregulates cytokinin (CK), a hormone that increases bud growth.  When a plant is decapitated, the auxin source is removed and bud growth and expression of Isopentenyl transferase (IPT) -  an enzyme involved in prenylation of adenines as cytokinins - are more likely to increase.  

Researchers have previously studied the interactions between auxin and cytokinin hormone signalling pathways in the *Arabidopsis thaliana* plant.  Analyses found very few differentially expressed genes.  We will analyze the same data and determine if we have obtained comparable results.

2.) Specifically, we would like to know if there are more or less genes that are differentially expressed than previously found. Once we have an answer to our question, we will discuss our findings.
  
3.) 

######Hypothesis:  

We will find the same number of genes that are differentially expressed between the B and E group if we run analyses at a .05 level.


4.) The data we will be using is from the *Arabidopsis thaliana* plant and obtained from NCBI GEO: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE59741.  The basis for our exploratory analyses is a research paper, "Cytokinin is required for escape but not release from auxin mediated apical dominance" (Muller et al, 2015).  Although a hypothesis was not clearly stated, the portion of the research paper that is of interest to us is located in the *Identification of auxin- and CK-responsive genes in buds* section.

Two sets of variables were tested.  One was Auxin-treated (A) vs Ethanol Control (E)  which we will refer to as the AvE group.  The other was  basal CK-treated plus Auxin-treated (B) vs Auxin-treated (A) control  which we will refer to as the BvA group.  A microarray study was conducted to compare the results of B with E (Diff) and a linear fitted model was used. At a .05 adjusted p-value, only 2 genes were found to be significantly differentially expressed.  At1g7540 (codes for cytokinin oxidase), and At2g33830 (associated with bud dormancy) were more highly expressed in the B group than in the E  group.

Additionally, genes with a fold-change of two or less in the E group but with a fold-change of two or more in the B group were determined to be significantly differentially expressed. 

####Anaylisis & Results

1.)  We will be using R to conduct the lmfit method used by Muller et al (2015). 

2.)

#####a.)

First, we used a text file and organized the data:

```{r}
library(limma)
library(affy)
pd = read.AnnotatedDataFrame("sample.txt",header=TRUE,sep=",",row.names=1)
mydata = ReadAffy(filenames=pd$filename,phenoData=pd,verbose = TRUE)
sampleNames(mydata)=row.names(pData(pd))
eset = rma(mydata)
expression_data = exprs(eset)

TS = factor(pd$treatment, levels = c("E","A", "B"))

exp_design = model.matrix(~ 0 +TS)
colnames(exp_design)=levels(TS)
```


#####b.)

Then we created a linear fit model of the expression data, compared the treatments to each other, created a matrix of the compared data, and applied empirical Bayes statitistics to the data:

```{r}
fit=lmFit(eset,exp_design)

contMatrix= makeContrasts(AvE=A-E, BvA=B-A, Diff= (B-A) - (E-A), levels= exp_design)

fit2 = contrasts.fit(fit, contMatrix)

diff_exp = eBayes(fit2)
```

#####c.)

Here, we set up a Venn diagram to view the differentially expressed genes at a .05 p-value:

```{r}

result = decideTests(diff_exp, p.value = 0.05)
vennDiagram(result)
```

There are two differentially expressed genes between B and E.

#####d.)

Then we used affy to retrieve information at the probeset level, specified top data with a p-value of .05, and chose to view the gene names of the compared treatments:
```{r}

f = "ftp://ftp.arabidopsis.org/Microarrays/Affymetrix/affy_ATH1_array_elements-2010-12-20.txt"
annots = read.delim(f, na.strings ="", fill=TRUE, header = T, sep="\t")
annots = annots[,c(1,5,6)]


N = dim(eset)[1]
top.A.AvE = toptable(diff_exp, coef=1, number = N, p.value = 0.05)
top.B.BvA = toptable(diff_exp, coef=2, number = N, p.value = 0.05)
top.Diff = toptable(diff_exp, coef = 3, number = N, p.value = 0.05)

#shows the gene names of compared treatments.
row.names(annots) = annots$array_element_name
top.B.BvA.merged = merge(annots, top.B.BvA, by = "row.names")
top.A.AvE.merged = merge(annots, top.A.AvE, by = "row.names")
top.Diff.merged = merge(annots, top.Diff, by = "row.names")

```

The only gene that is significantly differentially expressed in the Diff expression at a .05 level is , and it is a 

#####e.)

To show the genes from above that are shown, 

```{r}
genes.B.BvA = unique(top.B.BvA.merged$locus)
genes.Diff = unique(top.Diff.merged$locus)
genes.A.AvE = unique(top.A.AvE.merged$locus)

#shows the log fold changes of each compared treatment
logFC = top.Diff.merged[,'logFC']
logFC_AvE = top.A.AvE.merged[,'logFC']
logFC_BvE = top.B.BvA.merged[,'logFC']
```


3.)  Here is a histogram and scatter plot of the information we obtained through our analyses:

```{r}

hist(logFC)

plot(logFC)
plot(logFC_AvE)
plot(logFC_BvE)
```

####Limitations



####Discussion

####Works Cited
        
Müller, D., Waldie, T., Miyawaki, K., To, J. P., Melnyk, C. W., Kieber, J. J., Kakimoto, T., ...     Leyser, O. (January 01, 2015). Cytokinin is required for escape but not release from auxin         mediated apical dominance. The Plant Journal : for Cell and Molecular Biology, 82, 5, 874-86
  